---
title: "Untitled"
output: html_document
date: "2024-08-31"
---

# Data Structures

#### Vectors

A vector is simply a list of items that are of the same type.

To combine the list of items to a vector, use the `c()` function and separate the items by a comma.

In the example below, we create a vector variable called fruits, that combine strings:
```{r}
fruits <- c("banana", "apple", "peach")

fruits
```
```{r}
numbers <- 1:10
numbers
```
```{r}
# Vector with numerical decimals in a sequence
numbers1 <- 1.5:6.5
numbers1

# Vector with numerical decimals in a sequence where the last element is not used
numbers2 <- 1.5:6.3
numbers2
```
```{r}
# Vector of logical values
log_values <- c(TRUE, FALSE, TRUE, FALSE)

log_values
```
#### Vector Length
To find out how many items a vector has, use the `length()` function:
```{r}
fruits <- c("banana", "apple", "orange")

length(fruits)
```
#### Sort a Vector
To sort items in a vector alphabetically or numerically, use the `sort()` function:
```{r}
fruits <- c("banana", "apple", "orange", "mango", "lemon")
numbers <- c(13, 3, 5, 7, 20, 2)

sort(fruits)  # Sort a string
sort(numbers) # Sort numbers
```
#### Access Vectors
You can access the vector items by referring to its index number inside brackets `[]`. The first item has index 1, the second item has index 2, and so on:
```{r}
fruits <- c("banana", "apple", "orange")

# Access the first item (banana)
fruits[1]
```
```{r}
fruits <- c("banana", "apple", "orange", "mango", "lemon")

# Access the first and third item (banana and orange)
fruits[c(1, 3)]
```
```{r}
fruits <- c("banana", "apple", "orange", "mango", "lemon")

# Access all items except for the first item
fruits[c(-1)]
```
#### Change an Item
To change the value of a specific item, refer to the index number:
```{r}
fruits <- c("banana", "apple", "orange", "mango", "lemon")

# Change "banana" to "pear"
fruits[1] <- "pear"

# Print fruits
fruits
```
#### Repeat Vectors
To repeat vectors, use the `rep()` function:
```{r}
repeat_each <- rep(c(1,2,3), each = 3)

repeat_each
```
```{r}
repeat_times <- rep(c(1,2,3), times = 3)

repeat_times
```
```{r}
repeat_indepent <- rep(c(1,2,3), times = c(5,2,1))

repeat_indepent
```
Generating Sequenced Vectors
One of the examples on top, showed you how to create a vector with numerical values in a sequence with the `:` operator:
```{r}
numbers <- 1:10

numbers
```
To make bigger or smaller steps in a sequence, use the `seq()` function:
```{r}
numbers <- seq(from = 0, to = 100, by = 20)

numbers
```


