---
title: "R Notebook"
output: html_notebook
---
```{r}
library(ggplot2)
library(ggthemes)
library(lubridate)
library(dplyr)
library(tidyr)
library(tidyverse) # metapackage of all tidyverse packages
library(DT)
library(scales)
```

---

# Preparing the Data

### Explanation of R Code: Defining Colors

This code block stores specific color codes in a vector. These colors are often used in plots or visualizations.

- `colors = c(...)`: This line defines a vector named `colors`. The `c()` function combines multiple elements into a single vector.
- Each element is a color code in hexadecimal format (e.g., `#CC1011` is a red color).

The output of the code will display the contents of the `colors` vector.
```{r}
colors = c("#CC1011", "#665555", "#05a399", "#cfcaca", "#f5e840", "#0683c9", "#e075b0")
colors
```
### Explanation of R Code: Reading and Combining Uber Data

This code block reads Uber ride data for each month separately and then combines the data into a single dataset.

- **Reading Data:**
  - The `read.csv()` function is used to load CSV files for each month (April to September 2014) into R. Each file contains raw data related to Uber rides for the respective month.
  - The data is stored in variables named `apr`, `may`, `june`, `july`, `aug`, and `sept`, corresponding to the months from April to September.

- **Combining Data:**
  - The `rbind()` function is used to combine these monthly datasets row-wise, creating one large dataset named `data` that contains all the Uber rides from April to September 2014.
  
- **Displaying Data Dimensions:**
  - `dim(data)` returns the dimensions (number of rows and columns) of the combined dataset, which is printed using `cat()` to provide information about the size of the dataset.
```{r}
# Read the data for each month separately 
apr <- read.csv("/Users/tahaberkterekli/Documents/GitHub/R/projects/Uber Data Analysis/Uber-dataset/uber-raw-data-apr14.csv")
may <- read.csv("/Users/tahaberkterekli/Documents/GitHub/R/projects/Uber Data Analysis/Uber-dataset/uber-raw-data-may14.csv")
june <- read.csv("/Users/tahaberkterekli/Documents/GitHub/R/projects/Uber Data Analysis/Uber-dataset/uber-raw-data-jun14.csv")
july <- read.csv("/Users/tahaberkterekli/Documents/GitHub/R/projects/Uber Data Analysis/Uber-dataset/uber-raw-data-jul14.csv")
aug <- read.csv("/Users/tahaberkterekli/Documents/GitHub/R/projects/Uber Data Analysis/Uber-dataset/uber-raw-data-aug14.csv")
sept <- read.csv("/Users/tahaberkterekli/Documents/GitHub/R/projects/Uber Data Analysis/Uber-dataset/uber-raw-data-sep14.csv")

# Combine the data together 
data <- rbind(apr, may, june, july, aug, sept)
cat("The dimensions of the data are:", dim(data))
```
```{r}
# Print the first 6 rows of the data5
head(data)
```
### Explanation of R Code: Converting and Formatting Date and Time

This code block is responsible for converting the `Date.Time` column from a string to a date-time object and then extracting the time separately.

- **Converting `Date.Time` to POSIXct:**
  - `data$Date.Time <- as.POSIXct(data$Date.Time, format="%m/%d/%Y %H:%M:%S")`: This line converts the `Date.Time` column (currently stored as a string) into a POSIXct date-time object, which makes it easier to work with dates and times in R. The format used (`%m/%d/%Y %H:%M:%S`) corresponds to the structure of the date in the dataset (MM/DD/YYYY HH:MM:SS).

- **Extracting the Time:**
  - `data$Time <- format(as.POSIXct(data$Date.Time, format = "%m/%d/%Y %H:%M:%S"), format="%H:%M:%S")`: Here, the `format()` function is used to extract just the time (hours, minutes, and seconds) from the `Date.Time` column and store it in a new column named `Time`. This format is `%H:%M:%S`, which corresponds to hours, minutes, and seconds.

- **Converting `Date.Time` to POSIXct using `lubridate`:**
  - `data$Date.Time <- ymd_hms(data$Date.Time)`: This line uses the `ymd_hms()` function from the `lubridate` package to convert the `Date.Time` column to a standardized POSIXct format (with Year, Month, Day, Hour, Minute, Second).

This process ensures that both the date and time components are properly formatted for further analysis.
```{r}
data$Date.Time <- as.POSIXct(data$Date.Time, format="%m/%d/%Y %H:%M:%S")
data$Time <- format(as.POSIXct(data$Date.Time, format = "%m/%d/%Y %H:%M:%S"), format="%H:%M:%S")
data$Date.Time <- ymd_hms(data$Date.Time)
```
### Explanation of R Code: Extracting Day, Month, Year, and Day of the Week

This code block creates new columns in the dataset by extracting specific date components (day, month, year, and day of the week) from the `Date.Time` column.

- **Extracting Day:**
  - `data$day <- factor(day(data$Date.Time))`: The `day()` function (from the `lubridate` package) extracts the day of the month from `Date.Time`. The `factor()` function is then used to store the day as a factor, which is useful for categorical data analysis.

- **Extracting Month:**
  - `data$month <- factor(month(data$Date.Time, label=TRUE))`: The `month()` function extracts the month from `Date.Time`. Setting `label=TRUE` converts the numeric month (1-12) into a more readable label (e.g., Jan, Feb). The `factor()` function again ensures that months are treated as categorical variables.

- **Extracting Year:**
  - `data$year <- factor(year(data$Date.Time))`: The `year()` function extracts the year from `Date.Time`, and the `factor()` function converts it into a categorical variable. This could be useful if the data spans multiple years.

- **Extracting Day of the Week:**
  - `data$dayofweek <- factor(wday(data$Date.Time, label=TRUE))`: The `wday()` function extracts the day of the week from `Date.Time` (1 for Sunday, 7 for Saturday). By setting `label=TRUE`, it converts these numbers into weekday labels (e.g., Mon, Tue). The `factor()` function is used to handle it as categorical data.

This step helps in analyzing data based on time-related factors, like understanding trends across different days, months, or weekdays.
```{r}
data$day <- factor(day(data$Date.Time))
data$month <- factor(month(data$Date.Time, label=TRUE))
data$year <- factor(year(data$Date.Time))
data$dayofweek <- factor(wday(data$Date.Time, label=TRUE))
```
### Explanation of R Code: Extracting Hour, Minute, and Second

This code block adds new columns to the dataset by extracting specific time components (seconds, minutes, and hours) from the `Time` column.

- **Extracting Seconds:**
  - `data$second = factor(second(hms(data$Time)))`: 
    - The `hms()` function (from the `lubridate` package) converts the `Time` column, which is in "HH:MM:SS" format, into a time object.
    - The `second()` function extracts the second component (0-59) from this time object.
    - The `factor()` function converts the second values into a categorical variable, which can be useful for categorical analysis.

- **Extracting Minutes:**
  - `data$minute = factor(minute(hms(data$Time)))`: 
    - The `minute()` function extracts the minute component (0-59) from the `Time` column.
    - The `factor()` function converts it into a categorical variable for further analysis.

- **Extracting Hours:**
  - `data$hour = factor(hour(hms(data$Time)))`: 
    - The `hour()` function extracts the hour component (0-23) from the `Time` column.
    - Similarly, `factor()` is used to store hours as a categorical variable.

This step allows for analysis based on time variables, such as identifying trends based on specific hours, minutes, or seconds in the data.
```{r}
# Add Time variables as well 
data$second = factor(second(hms(data$Time)))
data$minute = factor(minute(hms(data$Time)))
data$hour = factor(hour(hms(data$Time)))
```
```{r}
head(data)
```

---

# Data Visualisation

### Explanation of R Code: Summarizing Data by Hour

This code block groups the dataset by the `hour` column and calculates the total number of entries (rides) for each hour. It then displays the summarized data in an interactive table.

- **Grouping Data by Hour:**
  - `data %>% group_by(hour)`: 
    - The `%>%` operator (pipe) is used to pass the `data` dataset into the next function.
    - `group_by(hour)` groups the data by the `hour` column, so that all rides occurring in the same hour are grouped together.

- **Summarizing the Total:**
  - `dplyr::summarize(Total = n())`: 
    - The `summarize()` function (from the `dplyr` package) creates a summary for each hour.
    - `n()` counts the number of rows in each group, representing the total number of rides for that specific hour.
    - This creates a new data frame, `hourly_data`, with two columns: `hour` and `Total`.

- **Displaying the Data:**
  - `datatable(hourly_data)`: This function (from the `DT` package) displays the `hourly_data` in an interactive table format, allowing for sorting, searching, and other interactive operations.

This code helps analyze how Uber rides are distributed across different hours of the day by summarizing the total ride counts for each hour.
```{r}
hourly_data <- data %>%
                    group_by(hour) %>%
                            dplyr::summarize(Total = n())

datatable(hourly_data)
```
### Explanation of R Code: Plotting Trips by Hour

This code block creates a bar chart that visualizes the total number of trips taken during each hour of the day using the `ggplot2` library.

- **Defining the Plot:**
  - `ggplot(hourly_data, aes(hour, Total))`: 
    - The `ggplot()` function initiates a plot using the `hourly_data` dataset. 
    - The `aes(hour, Total)` defines the aesthetics of the plot, with `hour` on the x-axis and `Total` on the y-axis, where `Total` represents the total number of trips for each hour.

- **Creating the Bar Chart:**
  - `geom_bar(stat="identity", fill="steelblue", color="red")`: 
    - `geom_bar()` creates a bar chart. The `stat="identity"` option means that the heights of the bars represent the values in the `Total` column, rather than being automatically calculated.
    - The bars are filled with a "steelblue" color and outlined with a red border, enhancing the chart's visual appeal.

- **Adding Title and Subtitle:**
  - `ggtitle("Trips Every Hour", subtitle = "aggregated today")`: 
    - This function adds a title ("Trips Every Hour") and a subtitle ("aggregated today") to the chart for context.

- **Customizing the Plot's Theme:**
  - `theme(legend.position = "none", plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5))`: 
    - `legend.position = "none"` removes the legend from the plot, as it is not needed.
    - `plot.title = element_text(hjust = 0.5)` and `plot.subtitle = element_text(hjust = 0.5)` center-align the title and subtitle, creating a balanced and visually appealing layout.

- **Formatting the Y-Axis:**
  - `scale_y_continuous(labels=comma)`:
    - This function modifies the y-axis by adding commas to the labels for better readability of large numbers (e.g., displaying "1,000" instead of "1000").

This code helps visualize how trips are distributed across different hours of the day, making it easier to spot trends or peak times for trip activity. The use of colors, title, and formatted labels makes the plot more informative and aesthetically pleasing.
```{r}
# Plot the data by hour
# Plot the data by hour
ggplot(hourly_data, aes(hour, Total)) + 
geom_bar(stat="identity", 
         fill="steelblue", 
         color="red") + 
ggtitle("Trips Every Hour", subtitle = "aggregated today") + 
theme(legend.position = "none", 
      plot.title = element_text(hjust = 0.5), 
      plot.subtitle = element_text(hjust = 0.5)) + 
scale_y_continuous(labels=comma)
```
### Explanation of R Code: Aggregating and Plotting Trips by Month and Hour

This code block aggregates the dataset by both month and hour, and then creates a bar chart to visualize the total number of trips for each combination of month and hour.

- **Aggregating the Data:**
  - `month_hour_data <- data %>% group_by(month, hour) %>% dplyr::summarize(Total = n())`: 
    - The `%>%` operator (pipe) is used to pass the `data` dataset into the next function.
    - `group_by(month, hour)` groups the data by both the `month` and `hour` columns, so that all trips occurring in the same month and hour are grouped together.
    - `dplyr::summarize(Total = n())` creates a summary for each group:
      - `n()` counts the number of rows (trips) in each group, representing the total number of trips for that specific month and hour.
    - This results in a new data frame, `month_hour_data`, with three columns: `month`, `hour`, and `Total`.

- **Creating the Bar Chart:**
  - `ggplot(month_hour_data, aes(hour, Total, fill=month)) +`: 
    - The `ggplot()` function initializes the plot using the `month_hour_data` dataset.
    - `aes(hour, Total, fill=month)` defines the aesthetics of the plot, with `hour` on the x-axis and `Total` on the y-axis. The bars are filled with different colors based on the `month` column.

  - `geom_bar(stat = "identity") +`: 
    - `geom_bar()` creates a bar chart. The `stat = "identity"` option specifies that the heights of the bars represent the values in the `Total` column.

  - `ggtitle("Trips by Hour and Month") +`: 
    - Adds a title ("Trips by Hour and Month") to the chart for context.

  - `scale_y_continuous(labels = comma)`: 
    - Formats the y-axis labels with commas for better readability of large numbers.

This code helps to analyze and visualize the distribution of trips across different hours of the day for each month. The use of colors for different months allows for easy comparison and trend identification across both time and seasonal variations.
```{r}
# Aggregate the data by month and hour
month_hour_data <- data %>% group_by(month, hour) %>% dplyr::summarize(Total =n())

ggplot(month_hour_data, aes(hour, Total, fill=month)) +
geom_bar(stat = "identity") +
ggtitle("Trips by Hour and Month") +
scale_y_continuous(labels = comma)
```
### Explanation of R Code: Aggregating Trips by Day

This code block aggregates the dataset by day and summarizes the total number of trips for each day.

- **Aggregating the Data:**
  - `day_data <- data %>% group_by(day) %>% dplyr::summarize(Trips = n())`: 
    - The `%>%` operator (pipe) is used to pass the `data` dataset into the next function.
    - `group_by(day)` groups the data by the `day` column, so that all trips occurring on the same day are grouped together.
    - `dplyr::summarize(Trips = n())` creates a summary for each group:
      - `n()` counts the number of rows (trips) in each group, representing the total number of trips for that specific day.
    - This results in a new data frame, `day_data`, with two columns: `day` and `Trips`.

- **Displaying the Data:**
  - `day_data`: 
    - This line prints the `day_data` data frame to the console, allowing you to view the aggregated total number of trips for each day.

This code helps analyze the total number of trips for each day, providing a clear summary of daily trip counts.
```{r}
day_data <- data %>% group_by(day) %>% dplyr::summarize(Trips = n())
day_data
```
### Explanation of R Code: Plotting Trips by Day of the Month

This code block creates a bar chart that visualizes the total number of trips for each day of the month using the `ggplot2` library.

- **Defining the Plot:**
  - `ggplot(day_data, aes(day, Trips)) +`: 
    - The `ggplot()` function initializes the plot using the `day_data` dataset.
    - `aes(day, Trips)` defines the aesthetics of the plot, with `day` on the x-axis and `Trips` on the y-axis, where `Trips` represents the total number of trips for each day.

- **Creating the Bar Chart:**
  - `geom_bar(stat = "identity", fill = "steelblue") +`: 
    - `geom_bar()` creates a bar chart. The `stat = "identity"` option means that the heights of the bars represent the values in the `Trips` column.
    - The bars are filled with a "steelblue" color, which provides a clear and consistent visual representation.

- **Adding Title:**
  - `ggtitle("Trips by day of the month") +`: 
    - Adds a title ("Trips by day of the month") to the chart for context.

- **Customizing the Plot's Theme:**
  - `theme(legend.position = "none") +`: 
    - `legend.position = "none"` removes the legend from the plot, as it is not needed for this type of visualization.

- **Formatting the Y-Axis:**
  - `scale_y_continuous(labels = comma)`: 
    - Formats the y-axis labels with commas for better readability of large numbers.

This code helps visualize the distribution of trips across different days of the month. The bar chart makes it easy to compare trip counts for each day and identify any patterns or trends.
```{r}
ggplot(day_data, aes(day, Trips)) +
geom_bar(stat = "identity", fill = "steelblue") +
ggtitle("Trips by day of the month") +
theme(legend.position = "none") + 
scale_y_continuous(labels = comma)
```

